Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-12-10 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_561f9}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The top right panel reveals a positive correlation between ipsilateral
and contralateral metastases. Extensive ipsilateral involvement likely
indicates a longer or faster disease progression. Additionally, it has
been hypothesized that bulky ipsilateral nodal disease may reroute
lymphatic drainage toward the contralateral side, potentially
facilitating contralateral metastasis.

\subsubsection{Interplay of Risk
Factors}\label{interplay-of-risk-factors}

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.56\% of advanced T-category tumors exhibit midline extension compared
to 6.13\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially accounts
for the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} isolates patients
with lateralized tumors that do not cross the midline, further
demonstrating these relationships. Among early T-category patients with
no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
patients) show involvement in contralateral level II. This proportion
increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
further to 39.68\% (25 of 63) for advanced T-category tumors with
ipsilateral levels II and III involved.

@@

We can collect the iteratively computed distributions for the midline
extension case to define the matrix over the states given all
time-steps, in analogy to equation~\ref{eq-lambda-matrix}:

@@

Using this, we can again write the joint of ipsi- and contralateral
involvement - now also for the case of mid-sagittal extension -
algebraically as before in equation~\ref{eq-bilateral-marginal-algebra}:

\[
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
\]

@@

The terms
\(P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)\)
and
\(P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)\)
are defined solely by sensitivity and specificity of the diagnostic
modality. Terms like these already appeared in the definition of the
observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
\(P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)\)
in the above equation is the crucial term that is supplied by a trained
model and its parameters \(\boldsymbol{\hat{\theta}}\).

It is possible to compute this \emph{posterior} probability of true
involvement not only for one fully defined state
\((\mathbf{X}^\text{i}, \mathbf{X}^\text{c})\), but also for
e.g.~individual LNLs: For example, the risk for involvement in the
contralateral level IV would be a marginalization over all ipsilateral
states \(\boldsymbol{\xi}^\text{i}_i\) and all contralateral states
\(\boldsymbol{\xi}^\text{c}_j\) where \(\xi^\text{c}_{j4}=1\). Formally:

\begin{equation}\phantomsection\label{eq-marg-over-posterior}{
P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
}\end{equation}

\section{Computational methods}\label{sec-methods}

@@

\subsection{Computing the Observed and Predicted Prevalence of
Involvement Patterns}\label{sec-prevalence}

We evaluate the model's ability to approximate the observed distribution
of lymphatic involvement patterns. We compare the prevalence of selected
involvement patterns in the data to the model's predicted prevalence,
given patient scenarios.

A ``scenario'' includes the patient's T-category \(\text{T}x\) and
whether the tumor extended over the mid-sagittal line,
i.e.~\(\epsilon=\texttt{True}\) or \(\epsilon=\texttt{False}\). An
involvement pattern specifies all ipsi- and contralateral LNLs' status
as ``healthy'', ``involved'', or ``masked'' (ignored).

For example, we may assess contralateral LNL II involvement prevalence
under early T-category (T0-T2) and no midline extension
(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
observed, with 27 exhibiting contralateral LNL II involvement -- a
prevalence of 7.12\%.

To visualize the data prevalence, we plot a \emph{beta posterior} over
the true prevalence \(q\). It is the result of a uniform beta prior over
\(q\), multiplied with the binomial likelihood for \(k\) out of \(n\)
patients, given \(q\). The resulting distribution has its maximum at
\(q=k / n\) and nicely captures the uncertainty in the observed cohort:
if only 3 out of 10 patients are observed with a specific pattern, the
distribution is much wider than if we observe 300 out of 1000.

@@

\[
P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
\]

In the enumator, we marginalize over all ipsi- and contralateral LNLs'
involvements, except for LNL II contralaterally. This is similar to the
marginalization in equation~\ref{eq-marg-over-posterior}, although we
are summing over different quantities. In the denominator, we can simply
insert the joint distribution over midline extension and diagnose time
\(P \left( \epsilon, t \right)\) marginalized over \(t\) using the early
T-category's time-prior.

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters. The bilateral model mostly
reproduces the ipsilateral spread parameter values reported in the
earlier publication on the unilateral model
\citep{ludwig_modelling_2023}. Any discrepancies may be due to
differences in the patient cohorts. Therefore, we omit the analysis of
the ipsilateral spread patterns and focus instead on the analysis of
contralateral involvement. Lastly, note the relatively small
contralateral spread parameters \(b^c_v\) compared to the ipsilateral
parameters \(b^i_v\). This accurately reflects the low prevalence of
contralateral lymph node involvement for lateralized tumors.

@@

\caption{}\label{T_d0076}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

The framework models the joint probability distribution of midline
extension and ipsi- and contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
This is visualized in figure~\ref{fig-model-algebra-equation}, which
represents the calculation defined in
equation~\ref{eq-bilateral-marginal-algebra}. To simplify
interpretation, the example focuses only on LNLs II, III, and IV,
reducing the state space to \(2^3 = 16\) possible states per side, and
\(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII are
excluded, along with their spread parameters, while remaining parameters
are set to their mean values from table~\ref{tbl-midline-params}.

@@

\caption{\label{fig-model-algebra-equation}Visual representation of
equation~\ref{eq-bilateral-marginal-algebra} for midline extension. The
left and right matrices represent the time evolution of hidden states
for the ipsi- and contralateral necks, respectively. The right matrices
distinguish between the cases of no midline extension (top) and midline
extension (bottom). The central diagonal matrix shows the time-prior for
late T-category tumors. This computation yields the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
visualized in figure~\ref{fig-model-state-dist}.}

@@

The left matrix in figure~\ref{fig-model-algebra-equation} shows the
time evolution of the probability distribution over the ipsilateral
involvement states, starting from the healthy state \([0,0,0]\). The two
right matrices show the contralateral state evolution, distinguishing
between the cases of midline extension and no midline extension. At
\(t=0\), the contralateral neck begins in the healthy state \([0,0,0]\)
without midline extension. The central matrix shows the time prior for
late T-category tumors. The matrix multiplication results in the joint
distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\),
visualized in figure~\ref{fig-model-state-dist}.

This joint distribution is presented as two heatmaps, corresponding to
the two states of midline extension. The most likely state involves a
lateralized tumor with ipsilateral level II involvement and no
contralateral involvement, having a probability of approximately 25\%.
The next most probable state involves ipsilateral levels II and III
without contralateral involvement. The most likely state with
contralateral involvement corresponds to tumors with midline extension,
showing involvement of contralateral level II and ipsilateral levels II
and III.

@@

\caption{\label{fig-model-state-dist}The full distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
over ipsi- and contralateral states and midline extension for late
T-category tumors. This three-dimensional distribution is shown as two
separate heatmaps for the binary midline extension variable
\(\epsilon\). These matrices are the result of
equation~\ref{eq-bilateral-marginal-algebra}, visualized in
figure~\ref{fig-model-algebra-equation}.}

@@

In figure~\ref{fig-model-prevalences-overall}, we compare the prevalence
of contralateral involvement for LNLs II, III, and IV across four
scenarios: combinations of early and late T-category, and lateralized
versus midline-extending tumors.

@@

\caption{\label{fig-model-prevalences-overall}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences, shown for the
contralateral LNLs II (blue), III (orange), and IV (green). The top row
shows scenarios with early T-category tumors, the bottom row for late
T-category ones. The left column depicts scenarios where the primary
tumor is clearly lateralized, the right column scenarios of tumors
extending over the mid-sagittal line. This figure illustrates the
model's ability to describe the prevalence of involvement for different
combinations of the risk factors T-category and midline extension.}

@@

Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
ability to account for key risk factors, such as T-category and midline
extension. Consistent with the data, the model predicts that the
prevalence of contralateral LNL II involvement increases from 7.1\% for
early T-category lateralized tumors to 39.17\% for advanced T-category
tumors that cross the midline. Similarly, contralateral LNL III
involvement rises from around 1.55\% for early T-category lateralized
tumors to nearly 14.25\% for advanced T-category midline-extending
tumors.

@@

Figure~\ref{fig-model-prevalences-upstream} highlights the influence of
upstream LNL II involvement on contralateral LNL III metastasis. The
contralateral LNL III rarely harbors metastases when its upstream LNL II
is healthy, a correlation well-captured by the model. However, for early
T-category tumors extending over the midline, the model slightly
underestimates the observed prevalence, likely due to the relative
rarity of such cases, resulting in broader uncertainty in the observed
prevalence estimates.

@@

\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
computed and observed prevalences for scenarios that illustrate the
model's capability of accounting for the correlation between ipsi- and
contralateral involvement. We show three scenarios where we consider the
joint involvement of contralateral LNL II together with different
ipsilateral involvements: 1) the ipsilateral neck shows no involvement
in blue (LNLs I to V are healthy, LNL VII is unspecified because data on
it is missing for some patients), 2) where ipsilateral LNL II is
involved in orange (LNLs I, III, IV, and V are healthy), and 3) where
ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V
are healthy). These two scenarios are plotted for all combinations of
T-category (early in top row, advanced in bottom row) and tumor
lateralization (lateralized in left column, extending over mid-sagittal
line in the right column).}

@@

For clinical applications, the model needs to estimate the risk of
occult metastases in clinically negative LNLs based on a patient's
diagnosis. The diagnosis includes the T-category, tumor lateralization
(whether it extends over the mid-sagittal plane), and the clinically
detected involvement of LNLs based on imaging and possibly fine needle
aspiration (FNA).

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_risks.pdf}}

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult involvement in contralateral LNL II (left), III (middle), and IV
(right), shown for various combinations of T-category, tumor
lateralization, and clinical LNL diagnoses. All LNLs not explicitly
mentioned in the legend, including the LNL for which the risk of occult
disease was computed, were assumed to be clinically negative
(specificity 76\%, sensitivity 81\%).}

@@

The predicted risk of occult disease in contralateral LNL II is shown in
the left panel of figure~\ref{fig-model-risks}. Tumor lateralization is
the strongest determinant of risk. For example:

\begin{itemize}
\tightlist
\item
  A patient with a lateralized early T-category tumor and ipsilateral
  LNL II involvement has a predicted risk of 1.55\% for occult
  contralateral LNL II disease (green histogram).\\
\item
  If the same tumor extends over the midline, the risk increases to
  7.58\% (orange histogram).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  An early T-category tumor crossing the midline with ipsilateral
  involvement of LNLs II and III has a 9.06\% risk of contralateral LNL
  II disease (red histogram).\\
\item
  For the same scenario with an advanced T-category tumor, the risk
  rises to 11.25\% (purple histogram).
\end{itemize}

The degree of ipsilateral involvement also influences risk. For a
midline-extending early T-category tumor, a clinically N0 ipsilateral
neck predicts a risk of 7.58\%, compared to 9.06\% when LNLs II and III
are involved.

In summary, midline extension is the primary risk factor for
contralateral LNL II involvement, but advanced T-category and extensive
ipsilateral involvement also contribute significantly.

@@

For contralateral LNL V, the risk is negligible unless there is severe
contralateral involvement, including LNL IV, confirmed by FNA. In such
extreme cases, the risk increases to 5.49\%.

@@

Despite its relatively few parameters, the model is highly
interpretable, with each parameter having a clear, intuitive
explanation. At the same time, it effectively describes the observed
data on ipsilateral and contralateral nodal involvement. To our
knowledge, this represents the most comprehensive model of lymphatic
tumor progression in oropharyngeal SCC, surpassing prior efforts, which
were conceptually different, limited in scope, or not trained on real
patient data \citep{benson_markov_2006, jung_development_2017}. The
underlying dataset and code are publicly available, supporting
reproducibility and further development.

>>>>

\caption{}\label{T_c6cfe}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The top right panel reveals a positive correlation between ipsilateral
and contralateral metastases. Extensive ipsilateral involvement likely
indicates a longer or faster disease progression. Additionally, it has
been hypothesized that bulky ipsilateral nodal disease may reroute
lymphatic drainage toward the contralateral side, potentially increasing
the probability of contralateral metastasis.

\subsubsection{Correlation of Risk
Factors}\label{correlation-of-risk-factors}

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.6\% of advanced T-category tumors exhibit midline extension compared
to 6.1\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially explains
the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} considers only
patients with lateralized tumors that do not cross the midline. Among
early T-category patients with no ipsilateral nodal involvement (levels
I-V), only 1.2\% (1 of 86 patients) show involvement in contralateral
level II. This proportion increases to 8.8\% (24 of 272) if ipsilateral
level II is involved, to 15.7\% (14 of 89) if ipsilateral levels II and
III are involved, and further to 22.2\% (12 of 54) for advanced
T-category tumors with ipsilateral levels II and III involved.

@@

We can collect the iteratively computed distributions for the midline
extension case to define the matrix over the states given all
time-steps, as in equation~\ref{eq-lambda-matrix}:

@@

In analogy to equation~\ref{eq-bilateral-marginal-algebra}, we can now
write the joint distribution of ipsi- and contralateral involvement and
midline extension algebraically:

\begin{equation}\phantomsection\label{eq-midline-marginal-algebra}{
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
}\end{equation}

@@

The terms
\(P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)\)
and
\(P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)\)
are defined solely by sensitivity and specificity of the diagnostic
modality. These terms already appeared in the definition of the
observation matrx in equation~\ref{eq-obs-matrix}. The \emph{prior}
\(P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)\)
in the above equation is the crucial term that is supplied by a trained
model and its parameters \(\boldsymbol{\hat{\theta}}\).

It is possible to compute this \emph{posterior} probability of true
involvement not only for one fully defined state
\((\mathbf{X}^\text{i}, \mathbf{X}^\text{c})\), but also for
e.g.~individual LNLs: For example, the risk for involvement in the
contralateral level IV would be a marginalization over all combination
of ipsi- states \(\boldsymbol{\xi}^\text{i}_i\) contralateral states
\(\boldsymbol{\xi}^\text{c}_j\) where \(\xi^\text{c}_{j4}=1\). Formally:

\begin{equation}\phantomsection\label{eq-marg-over-posterior}{
P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
}\end{equation}

\section{Computational Methods}\label{sec-methods}

@@

\subsection{Comparing the Observed and Predicted Prevalence of
Involvement Patterns}\label{sec-prevalence}

We evaluate the model's ability to describe the observed frequencies of
lymphatic involvement patterns. We compare the prevalence of selected
involvement patterns in the data to the model's predicted prevalence,
given patient scenarios. A ``scenario'' includes the patient's
T-category \(\text{T}x\) and whether the tumor extended over the
mid-sagittal plane, i.e.~\(\epsilon=\texttt{True}\) or
\(\epsilon=\texttt{False}\). An involvement pattern specifies all ipsi-
and contralateral LNLs' status as ``healthy'', ``involved'', or
``masked'' (ignored).

For example, in figure~\ref{fig-model-prevalences-overall} (top left
panel) we assess contralateral LNL II involvement prevalence for early
T-category (T0-T2) and no midline extension
(\(\epsilon=\texttt{False}\)). In the data, \(n=\) 379 such patients
were observed, with \(k=\) 27 exhibiting contralateral LNL II
involvement -- an observed prevalence of \(q=\) 7.1\%. To visualize the
data prevalence, we plot a \emph{beta posterior} over \(q\) -- with a
uniform beta prior -- multiplied with the binomial likelihood for \(k\)
out of \(n\) patients, given \(q\). The resulting distribution has its
maximum at \(q=k / n\) and nicely captures the statistical uncertainty
in the observed cohort: In the top right panel of
figure~\ref{fig-model-prevalences-overall}, we consider the case of
early T-category tumors extending over the midline. The dataset contains
only 29 such patients, out of which 6 had contralateral LNL II
involvement. Consequently, the beta distribution over the observed
prevalence is much wider.

@@

\[
\begin{multlined}
P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) \\ = \frac{ \sum_k \sum_{\ell \, : \, \xi_{\ell 2}=1}  P \left( \mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{ \sum_k \sum_\ell P \left(\mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
\end{multlined}
\]

In the enumator, we marginalize over all combinations of states in both
sides of the neck where the contralateral LNL II is involved. This is
similar to the marginalization in equation~\ref{eq-marg-over-posterior},
although we are summing over different quantities. In the denominator,
we simply sum out all LNL involvement, leaving only the joint
distribution over midline extension and diagnose time
\(P \left( \epsilon, t \right)\) marginalized over \(t\) using the early
T-category's time-prior.

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters. The bilateral model mostly
reproduces the ipsilateral spread parameter values reported in the
earlier publication on the unilateral model
\citep{ludwig_modelling_2023}. Any discrepancies may be due to
differences in the patient cohorts. Therefore, we omit the analysis of
the ipsilateral spread patterns and focus instead on the analysis of
contralateral involvement. The small contralateral spread parameters
\(b^c_v\) compared to the ipsilateral parameters \(b^i_v\) adequately
reflect the low prevalence of contralateral lymph node involvement for
lateralized tumors. The mixing parameter \(\alpha=\) 0.3 describes that
the probability of contralateral spread is higher for tumors extending
over the mid-sagittal plane, but still lower than the probability for
ipsilateral spread.

The high value of \(t_{23}\) (0.1) compared to the value of \(b_3\)
(0.3) shows that the ipsilateral LNL III is rarely involved without the
upstream involvement of LNL II.

@@

\caption{}\label{T_da35b}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

The framework models the joint probability distribution of midline
extension and ipsi- and contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
This is visualized in figure~\ref{fig-model-algebra-equation}, which
represents the calculation defined in
equation~\ref{eq-midline-marginal-algebra}. To simplify the
interpretation, the example focuses only on LNLs II, III, and IV,
reducing the state space to \(2^3 = 8\) possible states per side, and
\(2 \times 8 \times 8 = 128\) states in total. LNLs I, V, and VII are
excluded, along with their spread parameters, while remaining parameters
are set to their mean values from table~\ref{tbl-midline-params}.

@@

\caption{\label{fig-model-algebra-equation}Visual representation of
equation~\ref{eq-midline-marginal-algebra}. The left and right matrices
represent the time evolution of hidden states for the ipsi- and
contralateral necks, respectively. The right matrices distinguish
between the cases of no midline extension (top) and midline extension
(bottom). The central diagonal matrix shows the time-prior for late
T-category tumors. This computation yields the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
visualized in figure~\ref{fig-model-state-dist}.}

@@

The left matrix in figure~\ref{fig-model-algebra-equation} shows the
time evolution of the probability distribution over the ipsilateral
involvement states, starting from the healthy state \([0,0,0]\). The two
right matrices show the contralateral state evolution, distinguishing
between the cases of midline extension and no midline extension. At
\(t=0\), the contralateral neck begins in the healthy state \([0,0,0]\)
without midline extension. The central matrix shows the time prior for
late T-category tumors. The matrix multiplication results in the joint
distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\),
visualized in figure~\ref{fig-model-state-dist}.

This joint distribution is presented as two heatmaps, corresponding to
the two states of midline extension. The most likely state is a
lateralized tumor with ipsilateral level II involvement and no
contralateral involvement, having a probability of approximately 25\%.
The next most probable state is a lateralized tumor with ipsilateral
levels II and III involved, but without contralateral involvement. The
most likely state with contralateral involvement corresponds to tumors
with midline extension, showing involvement of contralateral level II
and ipsilateral levels II and III.

@@

\caption{\label{fig-model-state-dist}The joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) over ipsi-
and contralateral states and midline extension for late T-category
tumors. The distribution is shown as two separate heatmaps for the two
binary values of the midline extension variable \(\epsilon\). These
matrices are the result of equation~\ref{eq-midline-marginal-algebra},
visualized in figure~\ref{fig-model-algebra-equation}.}

@@

In figure~\ref{fig-model-prevalences-overall}, we compare the prevalence
of contralateral involvement for LNLs II, III, and IV.

@@

\caption{\label{fig-model-prevalences-overall}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences, shown for the
contralateral LNLs II (blue), III (orange), and IV (green). The top row
shows scenarios with early T-category tumors, the bottom row for late
T-category tumors. The left column depicts scenarios where the primary
tumor is clearly lateralized, the right column scenarios of tumors
extending over the mid-sagittal plane. This figure illustrates the
model's ability to describe the prevalence of involvement for different
combinations of the risk factors T-category and midline extension.}

@@

Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
ability to account for the risk factors T-category and midline
extension. Consistent with the data, the model predicts that the
prevalence of contralateral LNL II involvement increases from 7.1\% for
early T-category lateralized tumors to 39.2\% for advanced T-category
tumors that cross the midline. Similarly, contralateral LNL III
involvement rises from around 1.5\% for early T-category lateralized
tumors to nearly 14.2\% for advanced T-category midline-extending
tumors.

@@

Figure~\ref{fig-model-prevalences-upstream} highlights the influence of
upstream LNL II involvement on contralateral LNL III metastasis. The
contralateral LNL III rarely harbors metastases when its upstream LNL II
is healthy, a correlation well-captured by the model. Out of 164
patients with an advanced tumor crossing the midline and involvement in
the upstream LNL II, 19 had involvement in the contralateral LNL III. In
contrast, out of 379 patients with a clearly lateralized early
T-category tumor and no upstream involvement, only 1 showed
contralateral LNL III involvement. This is well captured by our model.

@@

\caption{\label{fig-model-prevalences-with-ipsi}Comparison of the
computed and observed prevalences for scenarios that illustrate the
model's capability of accounting for the correlation between ipsi- and
contralateral involvement. We show three scenarios where we consider the
joint involvement of contralateral LNL II together with different
ipsilateral involvements: 1) the ipsilateral neck shows no involvement
in green (LNLs I to V are healthy, LNL VII is unspecified because data
on it is missing for some patients), 2) where ipsilateral LNL II is
involved in orange (LNLs I, III, IV, and V are healthy), and 3) where
ipsilateral LNLs II and III are involved in red (LNLs I, IV, and V are
healthy). These three scenarios are plotted for all combinations of
T-category (early in top row, advanced in bottom row) and tumor
lateralization (lateralized in left column, extending over mid-sagittal
plane in the right column).}

@@

For clinical applications, the model needs to estimate the risk of
occult metastases in clinically negative LNLs based on a patient's
diagnosis. The diagnosis includes the T-category, tumor lateralization,
and the clinically detected involvement of LNLs based on imaging and
possibly fine needle aspiration (FNA).

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/figures/fig_model_risks.pdf}

@@

\caption{\label{fig-model-risks}Histograms over the predicted risk of
occult involvement in contralateral LNL II (top), III (middle), and IV
(bottom), shown for various combinations of T-category, tumor
lateralization, and clinical LNL diagnoses. All LNLs not explicitly
mentioned in the legend, including the LNL for which the risk of occult
disease was computed, were assumed to be clinically negative
(specificity 76\%, sensitivity 81\%).}

@@

The predicted risk of occult disease in contralateral LNL II is shown in
the left panel of figure~\ref{fig-model-risks}. Tumor lateralization is
the strongest determinant of risk:

\begin{itemize}
\tightlist
\item
  A patient with a lateralized early T-category tumor and ipsilateral
  LNL II involvement has a predicted risk of 1.6\% for occult
  contralateral LNL II disease (green histogram).\\
\item
  For an early T-category tumor that extends over the midline, with
  ipsilateral LNL II involved, the risk increases to 7.6\% (orange
  histogram).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  An early T-category tumor crossing the midline with ipsilateral
  involvement of LNLs II and III has a 9.1\% risk of contralateral LNL
  II disease (red histogram).\\
\item
  For the same scenario but an advanced T-category tumor, the risk rises
  to 11.3\% (purple histogram).
\end{itemize}

The degree of ipsilateral involvement also influences risk. For a
midline-extending early T-category tumor, and a clinically N0
ipsilateral neck, the predicted risk is 7.6\%. This increases to 9.1\%
when LNLs II and III are involved.

In summary, midline extension is the primary risk factor for
contralateral LNL II involvement, but advanced T-category and extensive
ipsilateral involvement also contribute.

@@

For contralateral LNL V, the risk is very small unless there is severe
contralateral involvement, including LNL IV, confirmed by FNA. In the
extreme case of advanced T-category tumors extending over the
mid-sagittal plane, with all ipsilateral LNLs, as well as contralateral
LNLs II, III, and IV involved, the risk increases only to 5.5\%.

@@

The model is highly interpretable, with each parameter having a clear,
intuitive explanation. Despite its relatively few parameters, it
adequately describes the observed data on ipsilateral and contralateral
nodal involvement. To our knowledge, this represents the most
comprehensive model of lymphatic tumor progression in oropharyngeal SCC,
surpassing prior efforts, which were conceptually different, limited in
scope, or not trained on real patient data
\citep{benson_markov_2006, jung_development_2017}. The underlying
dataset and code are publicly available, supporting reproducibility and
further development.
